Synthesizing Quotient Lenses

### 2018-03-10 papers/QuotientLenses/main.tex
More edits

\lstset{ language=Caml, basicstyle=\linespread{0.8}\upshape\sffamily,
keywordstyle=\upshape\sffamily\color{dkblue}, keepspaces=true,
framexleftmargin=1ex, framexrightmargin=1ex, showstringspaces=false,
keepspaces=true,
commentstyle=\itshape\rmfamily,
emph={synth,project,perm,squash,normalize,using,ins,del,lens},
emphstyle=\upshape\sffamily\color{dkblue}, columns=fullflexible,
% BCP: I find this distracting:
% stringstyle=\sffamily\color{dkred},
mathescape, }

@@

\newcommand{\FINISH}[3]{\ifdraft\textcolor{#1}{[#2: #3]}\fi}
\newcommand{\bcp}[1]{\FINISH{dkred}{B}{#1}}
\newcommand{\BCP}[1]{\FINISH{dkred}{B}{\bf #1}}
\newcommand{\afm}[1]{\FINISH{dkgreen}{A}{#1}}
\newcommand{\dpw}[1]{\FINISH{dkblue}{D}{#1}} % Toronto Maple Leafs Blue :-)
\newcommand{\saz}[1]{\FINISH{orange}{SZ}{#1}}
\newcommand{\SAZ}[1]{\FINISH{orange}{SZ}{\bf #1}}
\newcommand{\ksf}[1]{\FINISH{teal}{K}{#1}}
\newcommand{\sam}[1]{\FINISH{dkpurple}{SM}{#1}}

@@

Amongst the most restrictive lenses are the \emph{bijective lenses}.  These
lenses demand that get and put functions form a bijection, meaning no
information is lost when one translates from one data format to another.  While
this strong guarantee can be useful in some contexts, it is more often the case
that some of the information in one or both of the data formats is superfluous.
For instance, one typically does not care to preserve the amount of
whitespace between characters.  In other circumstances, capitalization
or the order of fields in a record or items in a list may be unimportant. Many
data formats have other ad hoc rules as well. In bibliographies, for example,
recall that in \bibtex{}, an author can be specified as
``\cd{lastname, firstname}'' or as ``\cd{firstname lastname}''---it does not
matter which of the two formats is used.

{\em Quotient lenses} loosen many of the restrictions of bijective lenses by
allowing a programmer to define an equivalence relation on the source and
target formats so that the round-tripping lens laws hold modulo these
equivalence rather than modulo equality. Quotient lenses thus make it possible
to align many more data sources than would otherwise be possible, by ignoring
details that a programmer views as inessential. The only language we are aware
of that implements a general class of quotient lenses is {\em
Boomerang}~\cite{quotientlenses}\bcp{Is this the right citation for
Boomerang?  Or should it be \cite{boomerang}?}. While Boomerang's quotient
lenses enable a 
programmer to define a very general class of quotient lenses\bcp{read
  literally, that part of the sentence is nonsense: boomerang's lenses
  (themselves) do not enable the programmer to define lenses.  Of course,
  the reader can probably figure out what is meant, but that's a needless
  tax on their finite mental energy.}, these lenses to have two main
drawbacks.

@@

The following list summarizes the central contributions of our work.
\begin{enumerate}
  \item We introduce {\em Quotient Regular Expressions} (QREs),
  a compact, convenient notation for simultaneously defining a
  regular language modulo some equivalence relation and a canonizer
  for that relation (Section~\ref{QRE}).
  \item We define a language of {\em QRE lenses}.  This language is
  a subset of the full set of bijective quotient
  lenses whose types are given by QREs. Our main technical contribution
  is a proof that each QRE lens is functionally equivalent to a QRE lens in the
  canonical form of a bijective lens with ``canonizers at the edges''.
   (Section~\ref{QRE-lenses}).
  \item We reduce the problem of {\em synthesizing}
  QRE lenses to the previously studied problem of synthesizing bijective lenses
  between regular languages (Section~\ref{synth}).
  \item We extend the Boomerang {\em implementation} with QREs
  and QRE lens synthesis.  We demonstrate the utility of our
  implementation by using it to
  synthesize QRE lenses between data formats drawn from the
  {\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.

@@

\section{QRE Lenses by Example}
\label{sec:example}
\subsection{Bijective Lenses}
Consider a bidirectional transformation that converts \bibtex{} citation records
such as
\begin{verbatim}
@Book {Doe,
Author = "John Doe",
Title = {Generic Title}
}
\end{verbatim}
\noindent
into an equivalent EndNote format like the following
\begin{verbatim}
%0 Book
%F Doe
%A John Doe
%T Generic Title
\end{verbatim}
\noindent
and vice versa.

{\em Bijective lenses} are designed to define a bidirectional transformation
such as this, where the data formats can be matched in a one-to-one manner, in
this case by matching the label, author and title fields. Boomerang allows
programmers to define such bijecive transformations using combinators which
favour a compositional approach in which the programmer defines small lenses
which are then put together using the combinators to define larger and larger
lenses until the final lens is defined.

Each bijective lens is constructed from simple primitive combinators including:
\begin{itemize}
  \item \kw{del} $s$: \textit{delete} the constant string $s$ in the get
  direction; insert it in the put direction.
  \item \kw{ins} $s$: \textit{insert} the constant string $s$ in the get
  direction; delete it in the put direction.
  \item \kw{copy} $R$: copy the text matching the regular expression $R$ in
  both directions.
\end{itemize}
In addition, such primitives may be combined using the regular operators
Kleene star, alternation and concatenation, or functional composition of
lenses.  Figure~\ref{fig:example-lens} illustrates the use of these
combinators to transform \bibtex{} to EndNote (and vice versa).
\begin{figure}[t]
\begin{lstlisting}
let preamble : (lens in "@book{" $\Leftrightarrow$ "%0 Book\n%F ") =
del "@book{" . ins "%0 Book\n%F "

@@

Unfortunately, programming bijective lenses in Boomerang is often slow and
tediuos because of the fiddly constraints that Boomerang imposes on its
terms. making lens programming slow and tedious. For example, Boomerang
programmers often must rearrange the order of data items by recursively using
operators that \swap adjacent fields. Furthermore, the Boomerang type checker is
very strict, disallowing many programs because they contain ambiguity about how
certain data is transformed.

In our previous work in~\cite{optician}, we presented a solution to these
shortcomings of bijective lenses by developing a tool called {\em Optician}
which enables a programmer to synthesize a bijective lens from a pair of regular
expressions $(S, T)$ and a set of input-output example \cd{exs} pairs that
determines how the synthesized lens should behave on those examples. With {\em
Optician}, the programmer need only spend the effort of specifying the data
formats and the behaviour\bcp{Let's use American spellings: behavior} of the
lens on a few inputs in order to define a 
bijective lens. As a result defining bijective lenses becomes faster and less
error prone, since the programmer need only invoke the synthesis procedure
\lstinline{synth S <=> T using exs} in order to define a bijective lens rather
than manually derive the lens herself.

\subsection{Quotient Lenses}
Unfortunately, while our synthesis tool greatly simplifies the task of
programming bijective lenses, not all bidirectional tranformations are bijective
in nature. For instance, users are typically uninterested in preserving
whitespace between words or between fields.  The order of author and title
fields is also likely irrelevant and there may be equivalent ways of writing
the same name:  ``Doe, John'' vs ``John Doe.''

@@

One natural way to define lenses that will synchronize data in these formats
is to provide the following items:
\begin{enumerate}
  \item A pair of descriptions of the external source and target formats, here
  \bibtex{} and EndNote. Such descriptions may be given by regular
  expressions, for instance.
  \item A pair of \emph{canonizers} that map external source (and target)
  objects into internal, canonical forms.  Such canonizers will take
  care of stripping away irrelevant information and otherwise normalizing
  differences between equivalent representations of the same logical data.
  When given an internal representation, such canonizers must also operate
  in the reverse direction, \emph{choosing} an element in the external format
  to represent the canonical internal element.
  \item A pair of descriptions of the internal source and
  target formats.
  \item A bijective lens to map back and forth between canonical elements
  of the internal source and target formats.
\end{enumerate}

@@

The first way in which our work simplifies Boomerang's quotient lenses is
that we merge the first three pieces of quotient lenses that we just described
into a single concept, {\em Quotient Regular Expression}, or QREs. A single QRE
enables a programmer to add annotations to the outer regular expression; these
annotations can then be used to infer an inner regular expression, as well as a
canonizer from the outer regular expression to the inner one. In the following
subsection, we describe the QRE combinators and use give an example of how a
programmer might define a QRE specification of \bibtex.

@@

Sometimes there are multiple disjoint representations of the same data.
In such situations, the \kw{squash} combinator creates a QRE that
allows external data to be in either format, and converts any
data in the first format to the second.
For instance, assume that
the \codefont{comma\_name} format describes ``Doe, John''
and that the \codefont{space\_name} format describes ``John Doe''
and \codefont{s\_to\_c} is a function from the first to the second.  In this case,
the following instance of \kw{squash} creates the desired canonizer.

@@

\begin{lstlisting}
let name = squash comma_name $\to$ space_name using s_to_c
\end{lstlisting}

One way to define the \codefont{s\_to\_c} canonization function used by squash
is simply to write it from scratch in some ordinary programming language. (Boomerang canonizers are written in OCaml.)
However, our system will also synthesize such functions automatically from pairs
of regular languages and examples---here, \codefont{s\_to\_c} is the
\codefont{get} direction of a lens that can be synthesized by:
%
\begin{lstlisting}
let l = synth comma_name $\Leftrightarrow$ space_name using {("Doe, John", "John Doe")}
let s_to_c = get l
\end{lstlisting}
%
\bcp{Three problems here. First, the idea of using half of a lens
as a canonizer comes out of nowhere.  The second is that this idea is a hack
(and always was).  Why bother defining / synthesizing a lens and then using only
half of it?  And the third is that, in the original quotient lens paper,
we used this trick on lenses that were not even bijective, which may confuse
people familiar with the earlier paper.}

\sam{OK I suppose it's a trick though I don't see an obvious solution. }

@@

(* synthesize a lens that maps comma representation to space representation *)
let l = synth comma_name $\Leftrightarrow$ space_name using {("Doe, John", "John Doe")}
let s_to_c = get l

@@

(* allow any permutation of fields interspersed with arbitrary whitespace *)
let bib_fields = perm (label, bib_author, bib_title) with (project ("," . wsp+) $\to$ ",\n")
let bibtex = "@book{" . bib_fields . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records. }
\label{fig:example-qre}
\end{figure}
\subsection{QRE Lenses and QRE Lens Synthesis}
\label{sec:examplesynth}
At this point we already have a tool for synthesizing bijective lenses from a
pair of regular expressions and a set of example input-output pairs, as well as
the language of QREs for defining equivalence relations on regular expressions.
An obvious next step in simplifying the quotient lens programming is thus to use
the bijective lens synthesis procedure as a subroutine for a quotient lens
synthesis procedure. This new procedure should take two QREs and a set
of example input-output pairs as input, compute the canonical data formats on
both sides of the lens from the pair of QREs, map the example input-output pairs
to their canonical representatives to get a set of examples in canonical form,
and then invoke the bijective lens synthesis procedure on the canonical data
formats and the new set of examples that are in canonical form.

Indeed this basic idea is what motivates our definition of {\em QRE Lenses}.
Intuitively, just like Boomerang's quotient lenses, our QRE lenses are bijective
lenses with ``canonizers at the edges''---Figure~\ref{fig:attheedges} presents
the architecture of each lens.
\begin{figure}[t]
\centering
\includegraphics[width=0.7\textwidth]{canonizers-outside}
\caption{Quotient Lens with ``Canonizers at the Edges''
\bcp{Maybe we should go a step further and label things like $K(c)$,
 $W(c)$, Canonizer(c), etc.  And probably we should display K as a subset of
 W.  I can make these changes later.)}}
\label{fig:attheedges}
\end{figure}
Every QRE lens $q_1$ has a type $q_1: c_1 \Leftrightarrow c_2$ where $c_1$ and $c_2$
are QREs. In the get direction, a QRE lens $q_1: c_1 \Leftrightarrow c_2$ uses the
source QRE $c_1$ to compute a unique representative for the data modulo the
equivalence relation defined by $c_1$ and then applies the $\get$ function of a
bijective lens $\ell_1$ to this representative. In the backward direction, $q_1$
operates similarly, but using the view QRE $c_2$ and the $\lput$ function of
$\ell_1$.

Because the QREs $c_1$ and $c_2$ determine the internal formats for data after
canonization, and because the algorithm for synthesizing bijective lenses is
directed by these formats, $c_1$ and $c_2$ are all that is required to
synthesize QRE lenses end-to-end. Indeed, we structured QRE lenses as bijective
lenses with canonizers at the edges, precisely to facilitate the use of this
synthesis algorithm.

@@

This technical investigation justifies replacing manually constructed lenses
using the QRE lens combinators with synthesized QRE lenses in normal form, thus
saving the programmer. For instance, after defining the \bibtex{} and EndNote
QREs and storing them in the variables \cd{bibtex} and \cd{endnote}
respectively, then all the code in Figure~\ref{fig:example-lens} may be
replaced by a single call to the synthesis prodedure:

\begin{lstlisting}
let bib_to_end : (lens in bibtex $\Leftrightarrow$ endnote) =
synth bibtex $\Leftrightarrow$ endnote using {(bib_example, end_example)}
\end{lstlisting}
\noindent Here, the generated  quotient lens synchronizes \cd{bibtex} and
\cd{endnote} formats, using \cd{bib_example} and \cd{end_example} (the two
concrete example strings given at the beginning of this section) to
disambiguate. In addition, and as we saw earlier with the definition of
\cd{s_to_c}, the synthesis procedure itself can be used to create lenses that
are in turn used to define other QREs.  The ability to intermix QRE
specification with QRE lens synthesis yields a powerful and flexible way of
creating bidirectional transformations.

\section{Quotient Regular Expressions}
\label{QRE}
\subsection{Syntax and Semantics of QREs}
Formally, the language of Quotient Regular Expressions (QREs) is given by the
following grammar:
\begin{align*}
c := \; &R \sep \project{R}{s} \sep \squash{c_1}{c_2}{f} \sep
\perm{c_1, \ldots, c_n}{c} \\
& | \; \normalize{R}{R'}{f} \sep c_1 \semicolon c_2 \sep c_1 \cdot c_2 \sep (c_1
\sep c_2) \sep c^*,
\end{align*}
where $c$ ranges over QREs, $R$ ranges over regular expressions, $f$ ranges over
functions between regular languages, and $s$ ranges over character strings.

Using the conventional notation that $\mathcal{L}(R)$ is the language accepted
by the regular expression $R$, then each QRE $c$ enables us to express
\begin{enumerate}
  \item a regular expression $W(c)$ (the ``whole'' of $c$),
  \item an equivalence relation $\eqrel{c}$ on $\mathcal{L}(W(c))$,
  \item a regular expression $K(c)$ (the ``kernel'' of $c$)
  such that $\mathcal{L}(K(c))$ forms a complete set of representatives for
  $\eqrel{c}$, and
  \item a ``canonizing'' function\bcp{Why is it called ``canonizer'' rather
    than ``canonize''??} $\canonizer(c):\mathcal{L}(W(c))
  \longrightarrow \mathcal{L}(K(c))$ which given any $w \in \mathcal{L}(W(c))$,
  computes $\canonizer(c)(w)$ as the unique $k$ in $\mathcal{L}(K(c))$ such that
  $k$ is equivalent to $w$ mod $\eqrel{c}$
\end{enumerate}
Figures~\ref{fig:wk} gives the inductive definitions for the whole and
kernel languages $W(c)$ and $K(c)$ of a QRE.
\begin{figure}[t]
\centering
\[
\begin{array}{l@{\quad}l@{\quad}l}

c & W(c) & K(c) \\ \hline
R & R & R \\
\project{R}{s} & R & s \\
\squash{c}{c_1}{f} & W(c) \sep W(c_1) & K(c_1) \\
\normalize{R}{R'}{f} & R & R' \\
c_1 \; ; \; c_2 & W(c_1) & K(c_2) \\
c_1 \cdot c_2 & W(c_1) \cdot W(c_2) & K(c_1) \cdot K(c_2) \\
c_1 \sep c_2 & W(c_1) \sep W(c_2) & K(c_1) \sep K(c_2) \\
c^* & W(c)^* & K(c)^* \\
\end{array}
\]
\[
\begin{array}{r@{\quad}l}
W( \perm{c_1, \ldots, c_n}{c} ) = &
\bigcup \limits_{\sigma \in S_n} W(c_{\sigma(1)}) \cdot W(c) \cdot \ldots \cdot
W(c) \cdot W(c_{\sigma(n)})\\
K( \perm{c_1, \ldots, c_n}{c} ) = & K(c_1) \cdot K(c) \cdot \ldots \cdot K(c)
\cdot K(c_n)
\end{array}
\]
\caption{Whole and Kernel Regular Expressions}
\label{fig:wk}
\end{figure}
The \textit{squash} and \textit{permutation} combinators give the two
interesting definitions. If $c = \squash{c_1}{c_2}{f}$, then the whole language
of $c$ is $W(c_1) \sep W(c_2)$. This is because the \textit{squash} combinator
takes the whole language $W(c_1)$ of $c_1$, the whole language $W(c_2)$ of $c_2$
and a function $f : \mathcal{L}(W(c_1)) \longrightarrow \mathcal{L}(W(c_2))$,
maps $\mathcal{L}(W(c_1))$ into $\mathcal{L}(W(c_2))$ using $f$ and then
canonizes $W(c_2)$ into $K(c_2)$ using $\canonizer(c_2)$.

For the $\perm{c_1, \ldots, c_n}{c}$ combinator, the whole language is the
union of languages of the form $W(c_{\sigma(1)}) \cdot W(c) \cdot \ldots \cdot
W(c) \cdot W(c_{\sigma(n)})$ for any permutation $\sigma$ in $S_n$, where
$S_n$ is the set of all permutations of the numbers $1$ to $n$. This is because
the \textit{permutation} combinator allows for the string to match any
permutation the $c_i$'s while also taking into account the separator $c$ in
between each of the $c_i$'s. The kernel of the \textit{permutation} combinator
is the language $K(c_1) \cdot K(c) \cdot \ldots \cdot K(c) \cdot K(c_n)$,
because the canonical permutation is the identity permuation, with each of the
parts of the string that match $c_i$ and $c$ canonized into $K(c_i)$ and $K(c)$
respectively.

>>>>

\lstset{ language=Caml, basicstyle=\linespread{0.8}\upshape\sffamily,
keywordstyle=\upshape\sffamily\color{dkblue}, keepspaces=true,
framexleftmargin=1ex, framexrightmargin=1ex, showstringspaces=false,
commentstyle=\itshape\rmfamily,
emph={synth,project,perm,squash,normalize,using,ins,del,lens},
emphstyle=\upshape\sffamily\color{dkblue}, columns=fullflexible,
% BCP: I find this distracting:
% stringstyle=\sffamily\color{dkred},
mathescape, }

@@

\newcommand{\FINISH}[3]{\ifdraft\textcolor{#1}{[#2: #3]}\fi}
\newcommand{\bcp}[1]{\FINISH{dkred}{B}{#1}}
\newcommand{\BCP}[1]{\FINISH{dkred}{B}{\bf #1}}
\newcommand{\afm}[1]{\FINISH{dkgreen}{A}{#1}}
\newcommand{\dpw}[1]{\FINISH{dkblue}{D}{#1}} % Toronto Maple Leafs Blue :-)
\newcommand{\saz}[1]{\FINISH{orange}{SZ}{#1}}
\newcommand{\SAZ}[1]{\FINISH{orange}{SZ}{#1}}
\newcommand{\ksf}[1]{\FINISH{teal}{K}{#1}}
\newcommand{\sam}[1]{\FINISH{dkpurple}{SM}{#1}}

@@

Amongst the most restrictive lenses are the \emph{bijective lenses}.
A bijective lens $\ell : S \Leftrightarrow T$ is simply a bijection from
a source set $S$ to a target set $T$:
\begin{equation}\label{bijectivelenslaws}
\ell.\get \; (\ell.\lput \; t) = t \text{, and } \ell.\lput \; (\ell.\get \; s)
= s
\end{equation}
Since these lenses demand that get and put functions form a bijection, then
no information is lost when one translates from one data format to another. 
While this strong guarantee can be useful in some contexts, it is more often
the case that some of the information in one or both of the data formats is
superfluous. For instance, one typically does not care to preserve the amount of
whitespace between characters.  In other circumstances, capitalization
or the order of fields in a record or items in a list may be unimportant. Many
data formats have other ad hoc rules as well. In bibliographies, for example,
recall that in \bibtex{}, an author can be specified as
``\cd{lastname, firstname}'' or as ``\cd{firstname lastname}''---it does not
matter which of the two formats is used.

{\em Quotient lenses}~\cite{quotientlenses} loosen many of the restrictions of
bijective string lenses by allowing a programmer to define an equivalence
relation on the source and target formats so that the round-tripping lens laws
hold modulo these equivalence rather than modulo equality. Quotient lenses thus
make it possible to align many more data sources than would otherwise be
possible, by ignoring details that a programmer views as inessential. The only
language we are aware of that implements a general class of quotient lenses is
{\em Boomerang}~\cite{boomerang}. While Boomerang enables a programmer to
define a very general class of quotient lenses, these lenses to have two main
drawbacks.

@@

The following list summarizes the central contributions of our work.
\begin{enumerate}
  \item We introduce {\em Quotient Regular Expressions} (QREs),
  a compact, convenient notation for simultaneously defining a
  regular language modulo some equivalence relation and a canonizer
  for that relation (Section~\ref{QRE}).
  \item We introduce {\em QRE lenses} which are lenses that map data specified
  using QREs.
  In addition, we will descibe a particular set of QRE lens combinators and
  prove that QRE lenses defined using these combinators can be transformed to
  a QRE lens that us defined using just two of these combinators; this is the
  main technical contribution of this paper.
  (Section~\ref{QRE-lenses}).
  \item We use this result to reduce the problem of {\em synthesizing} QRE
  lenses to the problem of synthesizing bijective string lenses which we studied
  in~\cite{optician}(Section~\ref{synth}).
  \item We extend the Boomerang {\em implementation} with QREs
  and QRE lens synthesis.  We demonstrate the utility of our implementation by
  using it to synthesize QRE lenses between data formats drawn from the
  {\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.

@@

\section{QRE Lenses by Example}
\label{sec:example}
\subsection{Bijective String Lenses}
Consider a bidirectional transformation that converts \bibtex{} citation records
such as
\begin{verbatim}
@Book {Doe,
Author = "John Doe",
Title = {Generic Title}
}
\end{verbatim}
\noindent
into an equivalent EndNote format like the following
\begin{verbatim}
%0 Book
%F Doe
%A John Doe
%T Generic Title
\end{verbatim}
\noindent
and vice versa.

Boomerang's bijective string lenses are designed to define a bidirectional
transformation such as this, where the data formats can be matched in a
one-to-one manner, in this case by matching the label, author and title fields.
Boomerang allows programmers to define such bijecive transformations using
combinators which favour a compositional approach in which the programmer
defines small lenses which are then put together using the combinators to
define larger and larger lenses until the final lens is defined.

Each bijective string lens is constructed from simple primitive combinators including:
\begin{itemize}
  \item \kw{del} $s$: \textit{delete} the constant string $s$ in the get
  direction; insert it in the put direction.
  \item \kw{ins} $s$: \textit{insert} the constant string $s$ in the get
  direction; delete it in the put direction.
  \item \kw{copy} $R$: copy the text matching the regular expression $R$ in
  both directions.
\end{itemize}
In addition, such primitives may be combined using the regular operators
Kleene star, alternation and concatenation, or functional composition of
lenses.  Figure~\ref{fig:example-lens} illustrates the use of these
combinators to transform \bibtex{} to EndNote (and vice versa).
\begin{figure}[t]
\begin{lstlisting}
let preamble : (lens in "@book{" $\Leftrightarrow$ "%0 Book\n%F ") =
del "@book{" . ins "%0 Book\n%F "

@@

Unfortunately, programming bijective string lenses in Boomerang is often slow and
tediuos because of the fiddly constraints that Boomerang imposes on its
terms. making lens programming slow and tedious. For example, Boomerang
programmers often must rearrange the order of data items by recursively using
operators that \swap adjacent fields. Furthermore, the Boomerang type checker is
very strict, disallowing many programs because they contain ambiguity about how
certain data is transformed.

In our previous work in~\cite{optician}, we presented a solution to these
shortcomings of bijective string lenses by developing a tool called {\em Optician}
which enables a programmer to synthesize a bijective string lens from a pair of regular
expressions $(S, T)$ and a set of input-output example \cd{exs} pairs that
determines how the synthesized lens should behave on those examples. With {\em
Optician}, the programmer need only spend the effort of specifying the data
formats and the behavior of the lens on a few inputs in order to define a
bijective string lens. As a result defining bijective string lenses becomes faster and less
error prone, since the programmer need only invoke the synthesis procedure
\lstinline{synth S <=> T using exs} in order to define a bijective string lens rather
than manually derive the lens herself.

\subsection{Quotient Lenses}
Unfortunately, while our synthesis tool greatly simplifies the task of
programming bijective string lenses, not all bidirectional tranformations are bijective
in nature. For instance, users are typically uninterested in preserving
whitespace between words or between fields.  The order of author and title
fields is also likely irrelevant and there may be equivalent ways of writing
the same name:  ``Doe, John'' vs ``John Doe.''

@@

One natural way to define lenses that will synchronize data in these formats
is to provide the following items:
\begin{enumerate}
  \item A pair of descriptions of the external source and target formats, here
  \bibtex{} and EndNote. Such descriptions may be given by regular
  expressions, for instance.
  \item A pair of \emph{canonizers} that map external source (and target)
  objects into internal, canonical forms.  Such canonizers will take
  care of stripping away irrelevant information and otherwise normalizing
  differences between equivalent representations of the same logical data.
  When given an internal representation, such canonizers must also operate
  in the reverse direction, \emph{choosing} an element in the external format
  to represent the canonical internal element.
  \item A pair of descriptions of the internal source and
  target formats.
  \item A bijective string lens to map back and forth between canonical elements
  of the internal source and target formats.
\end{enumerate}

@@

The first way in which our work simplifies Boomerang's quotient lenses is
that we merge the first three pieces of quotient lenses that we just described
into a single concept, {\em Quotient Regular Expression}, or QREs. A single QRE
enables a programmer to add annotations to the outer regular expression; these
annotations can then be used to infer an inner regular expression, as well as a
canonizer from the outer regular expression to the inner one. In the following
subsection, we describe a set of QRE combinators that we have integrated into
Boomerang and give an example of how a programmer might define a QRE
specification of \bibtex{} using these combinators.

@@

Sometimes there are multiple disjoint representations of the same data.
In such situations, the \kw{squash} combinator creates a QRE that
allows external data to be in either format, and converts any
data in the first format to the second.
For instance, assume that
the \codefont{comma\_name} format describes ``Doe, John''
and that the \codefont{space\_name} format describes ``John Doe''
and \codefont{c\_to\_s} is a function from the first to the second.  In this
case, the following instance of \kw{squash} creates the desired canonizer.

@@

\begin{lstlisting}
let name = squash comma_name $\to$ space_name using c_to_s
\end{lstlisting}

One way to define the \codefont{c\_to\_s} canonization function used by squash
is simply to write it from scratch in some ordinary programming language.
However, our system will also synthesize such functions automatically from pairs
of regular languages and examples---here, \codefont{c\_to\_s} is the
\codefont{get} direction of a lens that can be synthesized by:
%
\begin{lstlisting}
let l = synth comma_name $\Leftrightarrow$ space_name using {("Doe, John", "John Doe")}
let c_to_s = get l
\end{lstlisting}

@@

(* synthesize a lens that maps comma representation to space representation *)
let l = synth comma_name $\Leftrightarrow$ space_name using {("Doe, John", "John Doe")}
let c_to_s = get l

@@

(* allow any permutation of fields interspersed with arbitrary whitespace *)
let bib_fields = perm (label, bib_author, bib_title) with (project ("," . wsp+) $\to$ ",\n")
let bibtex = "@book{" . bib_fields . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records. }
\label{fig:example-qre}
\end{figure}
\subsection{QRE Lenses and QRE Lens Synthesis}
\label{sec:examplesynth}
At this point we already have a tool for synthesizing bijective string lenses from a
pair of regular expressions and a set of example input-output pairs, as well as
the language of QREs for defining equivalence relations on regular expressions.
An obvious next step in simplifying the quotient lens programming is thus to use
the bijective string lens synthesis procedure as a subroutine for a quotient lens
synthesis procedure. This new procedure should take two QREs and a set
of example input-output pairs as input, compute the canonical data formats on
both sides of the lens from the pair of QREs, map the example input-output pairs
to their canonical representatives to get a set of examples in canonical form,
and then invoke the bijective string lens synthesis procedure on the canonical data
formats and the new set of examples that are in canonical form.

Indeed this basic idea is what motivates our definition of {\em QRE Lenses}.
Intuitively, just like Boomerang's quotient lenses, our QRE lenses are bijective
lenses with ``canonizers at the edges''---Figure~\ref{fig:attheedges} presents
the architecture of each lens.
\begin{figure}[t]
\centering
\includegraphics[width=0.7\textwidth]{canonizers-outside}
\caption{QRE Lens with ``Canonizers at the Edges''}
\label{fig:attheedges}
\end{figure}
Every QRE lens $q_1$ has a type $q_1: Q_1 \Leftrightarrow Q_2$ where $Q_1$ and $Q_2$
are QREs. In the get direction, a QRE lens $q_1: Q_1 \Leftrightarrow Q_2$ uses the
source QRE $Q_1$ to compute a unique representative for the data modulo the
equivalence relation defined by $Q_1$ and then applies the $\get$ function of a
bijective string lens $\ell_1$ to this representative. In the backward direction, $q_1$
operates similarly, but using the view QRE $Q_2$ and the $\lput$ function of
$\ell_1$.

Because the QREs $Q_1$ and $Q_2$ determine the internal formats for data after
canonization, and because the algorithm for synthesizing bijective string lenses is
directed by these formats, $Q_1$ and $Q_2$ are all that is required to
synthesize QRE lenses end-to-end. Indeed, we structured QRE lenses as bijective
lenses with canonizers at the edges, precisely to facilitate the use of this
synthesis algorithm.

@@

This technical investigation justifies replacing manually constructed lenses
using the QRE lens combinators with synthesized QRE lenses with canonizers at
the edges. For instance, after defining the \bibtex{} and EndNote QREs and
storing them in the variables \cd{bibtex} and \cd{endnote} respectively, then
all the code in Figure~\ref{fig:example-lens} may be replaced by a single call
to the synthesis prodedure:

\begin{lstlisting}
let bib_to_end : (lens in bibtex $\Leftrightarrow$ endnote) =
synth bibtex $\Leftrightarrow$ endnote using {(bib_example, end_example)}
\end{lstlisting}
\noindent Here, the generated  quotient lens synchronizes \cd{bibtex} and
\cd{endnote} formats, using \cd{bib_example} and \cd{end_example} (the two
concrete example strings given at the beginning of this section) to
disambiguate. In addition, and as we saw earlier with the definition of
\cd{c_to_s}, the synthesis procedure itself can be used to create lenses that
are in turn used to define other QREs.  The ability to intermix QRE
specification with QRE lens synthesis yields a powerful and flexible way of
creating bidirectional transformations.

\section{Quotient Regular Expressions}
A Quotient Regular Expression (or QRE) is a regular expression $R$ augmented
with syntax that expresses an equivalence relation on the language of $R$. This
section describes in further detail the set of QRE combinators that we described
in \cref{sec:example}.
\label{QRE}
\subsection{Syntax and Semantics of QREs}
Formally, the language of Quotient Regular Expressions (QREs) is given by the
following grammar:
\begin{align*}
Q := \; &R \sep \project{R}{s} \sep \squash{Q_1}{Q_2}{f} \sep
\perm{Q_1, \ldots, Q_n}{Q} \\
& | \; \normalize{R}{R'}{f} \sep Q_1 \semicolon Q_2 \sep Q_1 \cdot Q_2 \sep (Q_1
\sep Q_2) \sep Q^*,
\end{align*}
where $Q$ ranges over QREs, $R$ ranges over regular expressions, $f$ ranges over
functions between regular languages, and $s$ ranges over character strings.

Using the conventional notation that $\mathcal{L}(R)$ is the language accepted
by the regular expression $R$, then each QRE $Q$ enables us to express
\begin{enumerate}
  \item a regular expression $W(Q)$ (the ``whole'' of $Q$),
  \item an equivalence relation $\eqrel{Q}$ on $\mathcal{L}(W(Q))$,
  \item a regular expression $K(Q)$ (the ``kernel'' of $Q$)
  such that $\mathcal{L}(K(Q))$ forms a complete set of representatives for
  $\eqrel{Q}$, and
  \item a ``canonizing'' function\bcp{Why is it called ``canonizer'' rather
  than ``canonize''??}\sam{because \canonizer(Q) is a function}
  $\canonizer(Q):\mathcal{L}(W(Q)) \longrightarrow \mathcal{L}(K(Q))$ which
  given any $w \in \mathcal{L}(W(Q))$, computes $\canonizer(Q)(w)$ as the
  unique $k$ in $\mathcal{L}(K(Q))$ such that $k$ is equivalent to $w$ mod
  $\eqrel{Q}$
\end{enumerate}
Figures~\ref{fig:wk} gives the inductive definitions for the whole and
kernel languages $W(Q)$ and $K(Q)$ of a QRE.
\begin{figure}[t]
\centering
\[
\begin{array}{l@{\quad}l@{\quad}l}

Q & W(Q) & K(Q) \\ \hline
R & R & R \\
\project{R}{s} & R & s \\
\squash{Q}{Q_1}{f} & W(Q) \sep W(Q_1) & K(Q_1) \\
\normalize{R}{R'}{f} & R & R' \\
Q_1 \; ; \; Q_2 & W(Q_1) & K(Q_2) \\
Q_1 \cdot Q_2 & W(Q_1) \cdot W(Q_2) & K(Q_1) \cdot K(Q_2) \\
Q_1 \sep Q_2 & W(Q_1) \sep W(Q_2) & K(Q_1) \sep K(Q_2) \\
Q^* & W(Q)^* & K(Q)^* \\
\end{array}
\]
\[
\begin{array}{r@{\quad}l}
W( \perm{Q_1, \ldots, Q_n}{Q} ) = &
\bigcup \limits_{\sigma \in S_n} W(c_{\sigma(1)}) \cdot W(Q) \cdot \ldots \cdot
W(Q) \cdot W(c_{\sigma(n)})\\
K( \perm{Q_1, \ldots, Q_n}{Q} ) = & K(Q_1) \cdot K(Q) \cdot \ldots \cdot K(Q)
\cdot K(Q_n)
\end{array}
\]
\caption{Whole and Kernel Regular Expressions}
\label{fig:wk}
\end{figure}
The \textit{squash} and \textit{permutation} combinators give the two
interesting definitions. If $Q = \squash{Q_1}{Q_2}{f}$, then the whole language
of $Q$ is $W(Q_1) \sep W(Q_2)$. This is because the \textit{squash} combinator
takes the whole language $W(Q_1)$ of $Q_1$, the whole language $W(Q_2)$ of $Q_2$
and a function $f : \mathcal{L}(W(Q_1)) \longrightarrow \mathcal{L}(W(Q_2))$,
maps $\mathcal{L}(W(Q_1))$ into $\mathcal{L}(W(Q_2))$ using $f$ and then
canonizes $W(Q_2)$ into $K(Q_2)$ using $\canonizer(Q_2)$.
